---
output:
  revealjs::revealjs_presentation:
    incremental: false
    self_contained: true
    center: true
    theme: default
    css: custom.css
    transition: slide
    slide_level: 1
    reveal_options:
      width: 90%
    includes:
      in_header: header.html
---

# Vectors

---

Start by making a vector with the numbers 1 through 26. 

Multiply the vector by 2.

Give the resulting vector names A through Z.

Hint: there is a built in vector called `LETTERS`

---

```{r}
x <- 1:26
x <- x * 2
names(x) <- LETTERS
x
```

---

Set the 10th and 13th elements of the vector you made in the last exercise to missing.

---

```{r}
x[c(10, 13)]<-NA
x
```

---

Select every other element, starting with the second, of the vector you're working with.  See the help for `seq` to help with this.

---

```{r}
x[seq(2, length(x), 2)]
```

---

# Data Frames

---

Make a data frame called `cats` that has this structure and values:

```
  coat weight likes_string
calico    2.1            1
 black    5.0            0
 tabby    3.2            1
```

---

```{r}
cats<-data.frame(coat=c("calico","black","tabby"), 
                 weight=c(2.1, 5, 3.2), 
                 likes_string=c(1,0,1))
```

---

Select the rows from your data frame where the cats' weight is greater than 3.

---

```{r}
cats[cats$weight > 3,]
```

---

What does each of the following do?

```
cats[1]
cats[[1]]
cats$coat
cats["coat"]
cats[1, 1]
cats[, 1]
cats[1,]
cats[-1,]
```

Hint: Use the function `typeof()` to examine what is returned in each case.

---

Add a column for `age` to `cats`.  Fill it in with whatever values you want.

---

```{r}
cats$age<-c(3,5,7)
```


# Reading Data

---

Read in the data (a csv file) from https://goo.gl/IFCKAm to a data frame called `gapminder`.  Do you need to make any adjustments to the defaults?

Look at the type of each column in the data you read in.  

View the observations for Belgium.

---

```{r, eval=FALSE}
gapminder<-read.csv("https://goo.gl/IFCKAm")
str(gapminder)
View(gapminder[gapminder$country=="Belgium",])
```

```{r,echo=FALSE}
gapminder<-read.csv("https://goo.gl/IFCKAm")
```

---

Read in the tab-delimited file (states.txt) at https://goo.gl/AwnS4R.  What changes do you need to make to read in tab-delimited data?  Are any other adjustments to the defaults warranted?

Change the column names to state, lat, and lon.  

Does anything look strange to you about this data?

---

```{r}
statedata<-read.csv("https://goo.gl/AwnS4R", sep='\t')
names(statedata)<-c("state", "lat","lon")
dim(statedata)
```

There are more than 50 observations.  Take a look at the data, and you'll see some unfamiliar codes.  These correspond to US territories.


---

# Loops and Conditionals

---

Use an if() statement to print a suitable message reporting whether there are any records from 2002 in the gapminder dataset. Now do the same for 2012.

Hint: use the `any` function

---
```{r, eval=FALSE}
if(any(gapminder$year == 2002)){
   print("Record(s) for the year 2002 found.")
}
```

or

```{r, eval=FALSE}
if(nrow(gapminder[(gapminder$year == 2002),]) >= 1){
   print("Record(s) for the year 2002 found.")
}
```

---

Write a script that loops through the `gapminder` data by continent and prints out whether the mean life expectancy is smaller or larger than 50 years.

---

```{r}
thresholdValue <- 50

for(iContinent in unique(gapminder$continent)) {
   tmp <- mean(subset(gapminder, continent==iContinent)$lifeExp)
   
   if(tmp < thresholdValue) {
       cat("Average Life Expectancy in", iContinent, "is less than", thresholdValue, "\n")
   } else {
       cat("Average Life Expectancy in", iContinent, "is greater than", thresholdValue, "\n")
   } # end if else condition
   
} # end for loop
```

---

Modify the script from the last exercise to loop over each country. This time print out whether the life expectancy is smaller than 50, between 50 and 70, or greater than 70.

---

```{r}
lowerThreshold <- 50
upperThreshold <- 70
 
for(iCountry in unique(gapminder$country)){
    tmp <- mean(subset(gapminder, country==iCountry)$lifeExp)
    
    if(tmp < lowerThreshold){
        cat("Average Life Expectancy in", iCountry, "is less than", lowerThreshold, "\n")
    }
    else if(tmp > lowerThreshold && tmp < upperThreshold){
        cat("Average Life Expectancy in", iCountry, "is between", lowerThreshold, "and", upperThreshold, "\n")
    }
    else{
        cat("Average Life Expectancy in", iCountry, "is greater than", upperThreshold, "\n")
    }
   
}
```


---

# `dplyr`

---

Using the `gapminder` data, write a single command (which can span multiple lines and includes pipes) that will produce a dataframe that has the African values for lifeExp, country and year, but not for other Continents.

---

```{r, eval=FALSE}
gapminder %>%
  filter(continent=="Africa") %>%
  select(year,country,lifeExp)
```

---

Calculate the average life expectancy per country. Which has the longest average life expectancy and which has the shortest average life expectancy?

---

```{r, eval=FALSE}
lifeExp_bycountry <- gapminder %>%
   group_by(country) %>%
   summarize(mean_lifeExp=mean(lifeExp))

lifeExp_bycountry %>% 
   filter(mean_lifeExp == min(mean_lifeExp) | mean_lifeExp == max(mean_lifeExp))

lifeExp_bycountry %>%
   arrange(mean_lifeExp) %>%
   head(1)

lifeExp_bycountry %>%
   arrange(desc(mean_lifeExp)) %>%
   head(1)
```

---

Calculate the average life expectancy in 2002 of 2 randomly selected countries for each continent. Then arrange the continent names in reverse order. Hint: Use the dplyr functions `arrange()` and `sample_n()`, they have similar syntax to other dplyr functions.

---
```{r, echo=FALSE, message=FALSE, warning=FALSE}
library(dplyr)
library(ggplot2)
```


```{r}
lifeExp_2countries_bycontinents <- gapminder %>%
   filter(year==2002) %>%
   group_by(continent) %>%
   sample_n(2) %>%
   summarize(mean_lifeExp=mean(lifeExp)) %>%
   arrange(desc(mean_lifeExp))
```

---

# `ggplot2`

---

Using the `gapminder` data, plot life expectancy vs. year.  Then color the points by continent.

---

```{r}
ggplot(data = gapminder, aes(x = year, y = lifeExp)) + geom_point()
```

---

```{r}
ggplot(data = gapminder, aes(x = year, y = lifeExp, color=continent)) +
  geom_point()
```

---

Starting with 

```{r, eval=FALSE}
ggplot(data = gapminder, aes(x = gdpPercap, y = lifeExp)) +
  geom_point() + scale_x_log10() + geom_smooth(method="lm", size=1.5)
```

1) Modify the color and size of the points on the point layer.

2) Modify your solution so that the points are now a different shape and are colored by continent with new trendlines.

---

```{r}
ggplot(data = gapminder, aes(x = gdpPercap, y = lifeExp)) +
 geom_point(size=3, color="orange") + scale_x_log10() +
 geom_smooth(method="lm", size=1.5)
```

---

```{r}
ggplot(data = gapminder, aes(x = gdpPercap, y = lifeExp, color = continent)) +
  geom_point(size=3, shape=17) + scale_x_log10() +
  geom_smooth(method="lm", size=1.5)
```


---

# Data Manipulation

---

Add a new column to `gapminder` that is the population in millions.

Make a new data frame, `gapminder2007` that just contains the observations for 2007.

---

```{r}
gapminder$popmillion <- gapminder$pop/1000000
gapminder2007 <- gapminder[gapminder$year==2007,]
```

---

Do a t-test using `gapminder2007` to see if life expectancy is different in Europe and Asia.

---

```{r}
t.test(gapminder2007$lifeExp[gapminder2007$continent=="Europe"],
       gapminder2007$lifeExp[gapminder2007$continent=="Asia"])
```

---




