---
output:
  revealjs::revealjs_presentation:
    incremental: false
    self_contained: true
    center: true
    theme: default
    css: custom.css
    transition: slide
    slide_level: 1
    includes:
      in_header: header.html
---

# Intro to R {data-background-color="#4E2A84" .white-font}

\
\


Christina Maimone \
christina.maimone@northwestern.edu\
Research Computing Services


---

## Why R 

## Things to Know 

## Differences from Stata/Other Programs 

## Ways to Run R 

## Sharing R Analysis 



# Why learn R? { data-background-color="#E4E0EE"} 
\


---

## It's free

---

## Customizable, extensible

There's something you want to do that's difficult or not available in Stata \
\
Many contributed libraries

---

## Full programming language

But still designed for and focused on data analysis

---

## Collaboration with colleagues

Be able to read and evaluate code and analysis from others

---

## Support for reproducible analysis

Methods of publishing code and results


# Things to Know {data-background-color="#E4E0EE"}

---

## R runs in memory

Mostly relevant for large data sets

---


## "R" refers to the language and the environment

---

## User contributed packages = possible mistakes, bugs, errors

But there are many well-established, trustworthy packages too

---

## Packages are installed separately for each version of R

---

## There are multiple ways to do most things...  

<h2 class="hcenter fragment">
but some ways are better than others</h2>


# Some Differences from Stata/Other Statistical Programs { data-background-color="#E4E0EE"}

---

## Datasets/Variables

* You can (and often do) have more than one dataset open in R
* Datasets, vectors, scalars, etc. are all objects stored with names 
* It's more common to have variables unattached to a dataset (from Stata: macros/scalars)
* No metadata on datasets (possible, but not done)

---

## R Function Syntax

Functions (operations on data/objects) have the form:

```{r, eval=FALSE}
functioname(unnamed_argument_value, named_argument_name=value)
```

Basic arithmetic and operations for indexing or subsetting objects (e.g. datasets, vectors, variables, lists) are different

---

## R functions can't change the value of a variable/dataset

You have to save the value/altered data returned from a function back to the variable/datasets

```{r, eval=FALSE}
mydataset <- read.csv("mydata.csv")
myvector <- sort(myvector)
```

---

## A lot of R functions are in packages/libraries that you have to install and load

---

## Comment character is `#`

`#` applies to everything after it on a line

No multi-line comments (use RStudio to comment/uncomment multiple lines at once)


---


# Ways to Run R {data-background-color="#E4E0EE"}


---

## Interactively

R Console

RStudio

Jupyter Notebooks

---

## Batch Jobs

`Rscript myfile.r`

---

## Call from other programs


# Where to Run R {data-background-color="#E4E0EE"} 


---

## On your own computer

---

## On a remote server

Interactively

In a web browser

---

## On a computing cluster


# Sharing Analysis from R {data-background-color="#E4E0EE"}

_In addition to the typical analysis workflow, you can also create files for local/offline use or publish them publicly using:_


# Shiny 

Create web applications 

https://shiny.rstudio.com/

# {data-background-image="https://shiny.rstudio.com/images/debugging/kmeans-showcase.gif" data-background-size=contain}


# RMarkdown

Write presentations (including this one), reports, notebooks

Can share via RPubs: http://rpubs.com/


# Plotly

Interactive, well-styled plots 

https://plot.ly/r/

# {data-background-image="https://cpsievert.github.io/plotly_book/images/gapminder-highlight-animation.gif" data-background-size=contain}

---

# Workshop Materials {data-background-color="#4E2A84" .white-font}

https://nuitrcs.github.io/rworkshops/intror_june2017/

or, shorter:

https://goo.gl/CdcMx8




